---
title: "Using the 'vader' R package for sentiment analysis"
author: "Megan Chiovaro, University of Connecticut, Department of Psychological Sciences, Center for the Ecological Study of Perception and Action"
date: "2/16/2021"
output: html_document
---

### "vader" Package for R

The 'vader' package for R is used to quantify a compound valence score, and the weighted percentages of positive, negative, and neutral words of a text. First used as a Python tool (https://github.com/cjhutto/vaderSentiment), the methods of vader have since been made into an R package with two primary functions: 'get_vader' and 'vader_df'. Today we will practice using both!

## Set up

In order to use the 'vader' package for R, we first need to install and load it.

```{r setup, include=FALSE}

# turn on caching: This saves the results from a chunk so that when you knit, it does not re-run the code chunk permitting all lines of code have remained the same. This saves on computation time!
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.lazy=FALSE)

# install packages
install.packages("vader")
install.packages("ggplot2")

# load packages
library(vader)
library(ggplot2)

```

## Read in data

To practice using 'vader,' let's create some fake data!

```{r data}

# create fake data to practice
data <- as.data.frame(c("I love the outdoors, it's beautiful.", 
                        "Today is a sad day.", 
                        "Pluto is no longer a planet."))

```

## Use the 'get_vader' function 

Using the 'get_vader' function will produce a compound valence score, and the weighted percentages of positive, negative, and neutral words in a single text.

```{r get_vader_word}

# try just printing the results for a single word so we can see the output
get_vader("hi")

# we can also save the results to an object for saving
hi_sentiment <- as.data.frame(get_vader("hi"))

```

We can see that the function produces 6 output metrics: word_scores, compund, pos, neu, neg, and but_count.

word_scores: a list of the individual word values in the text
compound: the overall valence score of the text
pos: the weighted average of positive words in the text
neu: the weighted average of neutral words in the text
neg: the weighted average of negative words in the text
but_count: the number of time "but" shows up in the text

```{r get_vader_sentence}

# run vader on the first sentence in our dataframe
get_vader("happy, good, terrible, sad")

```

Now we can see the list of scores for each word in the "sentence." There are four words, and the word_scores list has four values. 

The 'get_vader' function is quick to use on a single text, but what about when we have multiple pieces of text that we want to analyze?

## Use the 'vader_df' function 

This is where 'vader_df' comes in! This function allows you to run the sentiment analysis over a vector or within a column. Here, we will run it over out "data" that we made earlier.

```{r vader_df}

# run vader on the dataframe, hold results in a new dataframe, and view results
(vader_data_mult <- vader_df(data))

```

This produced a dataframe where each line holds the output metrics for each line of text that was in the original dataframe.

## Analyzing shifts in sentiment over time

Let's pretend we are talking to a little kid who has a fear of the dark, and their physician records the child's feelings about the dark at every visit.

```{r time-evol-sentiment}

# create some text data
evol_data <- as.data.frame(c("I am scared of the dark, there are monsters under the bed that are very bad. I hate them. They make me feel in danger.", 
                        "I do not like the dark, but it is less scary now.", 
                        "The dark is okay. I close my eyes and it is okay.",
                        "I like the dark. I can look at the stars outside.",
                        "I love the dark! It is fun and great to tell mystery stories!")) 

# run vader on the data
vader_child_data <- vader_df(evol_data)


# Plots in base R
plot(vader_child_data$compound, type = "l",  xlab = "Time", ylab = "Compound Sentiment")
plot(vader_child_data$pos, type = "l",  xlab = "Time", ylab = "Positive Sentiment")
plot(vader_child_data$neu, type = "l",  xlab = "Time", ylab = "Neutral Sentiment")
plot(vader_child_data$neg, type = "l",  xlab = "Time", ylab = "Negative Sentiment")

```

This can also be done in ggplot but the ggplot lesson is later in the semester, so I will not go over the code but I have left it here for those that are familiar!

```{r ggplot-sentiment}

# plot the time series of the compound, pos, neu, and neg values
ggplot() +
  
  # add compound sentiment
  geom_line(data = vader_child_data, aes(x = c(1:5), y = compound, color = "compound")) +
  
  # add positive sentiment
  geom_line(data = vader_child_data, aes(x = c(1:5), y = pos, color = "positive")) +
  
  # add neutral sentiment
  geom_line(data = vader_child_data, aes(x = c(1:5), y = neu, color = "neutral")) +
  
  # add negative sentiment
  geom_line(data = vader_child_data, aes(x = c(1:5), y = neg, color = "negative")) +
  
  # label axes
  xlab("Time") +
  ylab("Sentiment score")

```

We can see that as time goes on, the child generally increases the postivity of their statements and reduces their use of negative feelings toward the darkness!

## For good measure

"Luke, I am your sentiment analysis tool for R" - vader